---
title: "Locum"
output: github_document
---

The [`wrapr dot arrow pipe`](https://journal.r-project.org/archive/2018/RJ-2018-042/index.html) was designed
to both be an effective [`R`](https://www.r-project.org) function application pipe *and* also an experimental test-bed for
pipe effects.

Let's take a look at implementing a new effect from a user perspective.  The idea we want to implement is delayed evaluation through a 
collecting object we call a "locum" or stand-in.  The locum is intended to collect operations without executing them, for later use.  This is similar to a lambda or function abstraction.  The [code is now in the `wrapr` package](https://github.com/WinVector/wrapr/blob/master/R/locum.R), but could be implemented by any user as it uses only public or semi-public `wrapr` interfaces.

The `locum` was for a while a part of `wrapr`, but now it is user code in the file `locum.R`.

Let's start loading the `wrapr` package.

```{r}
library(wrapr)
source("locum.R")
```


We can use the `locum` to collect the operations, and then print them.

```{r}
y <- 4
p <- locum() %.>% 
  sin(.) %.>% 
  cos(.) %.>% 
  atan2(., y)

print(p)
```

We can now replace the `locum` with the value we want to apply the pipeline to.

```{r}
5 %.>% p
```

This yields the same answer as the following function application.

```{r}
atan2(cos(sin(5)), 4)
```

We can also add later intended arguments to the pipeline formatting.

```{r}
print(p, 'start' = 4)
```

We can do some fun things, such as combining `locum` pipelines.

```{r}
p1 <- locum() %.>% sin(.)
p2 <- locum() %.>% cos(.)

p12 <- p1 %.>% p2
p12
```

```{r}
4 %.>% p12
```

```{r}
cos(sin(4))
```



The idea is: `wrapr` dot arrow pipe is designed for expansion through the `apply_right`, `apply_left` `S3` interfaces, and the `apply_right_S4` `S4` interface.  And users can access the implementation through the `pipe_impl` function. [This example](https://github.com/WinVector/wrapr/blob/master/Examples/Locum/Locum.md) and [the formal article](https://journal.r-project.org/archive/2018/RJ-2018-042/index.html) should give users a good place to start. [`rquery`](https://github.com/WinVector/rquery), [`rqdatatable`](https://github.com/WinVector/rqdatatable), and [`cdata`](https://github.com/WinVector/cdata) already use the extension interfaces to implement interesting features.


