---
title: "Frame Tools"
author: "John Mount, Win-Vector LLC"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Frame Tools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

[`wrapr`](https://winvector.github.io/wrapr/) supplies a few tools for creating example 
`data.frame`s.  An important use case is: building the control table for `cdata::rowrecs_to_blocks()` and `cdata::blocks_to_rowrecs()` (example [here](https://winvector.github.io/cdata/articles/cdata.html)).

Lets see how to create an example data frame.  The idea is similar to that found in [`tibble::tribble()`](http://tibble.tidyverse.org/reference/tribble.html): for small tables a 
row oriented constructor can be quite legible, and avoids the cognitive load of taking a transpose.

For example we can create a typical `data.frame` as follows:

```{r}
d <- data.frame(
  names = c("a", "b", "c", "d"),
  x =     c(1,   2,   3,   4  ),
  y =     c(1,   4,   9,   16 ),
  stringsAsFactors = FALSE)

print(d)
```

Notice how the table is specified by columns (which is close to how `data.frame`s
are implemented), but printed by rows.  `utils::str()` and `tibble::glimpse()` both 
print by columns.

```{r}
str(d)
```

`wrapr` supplies the method [`draw_frame`](https://winvector.github.io/wrapr/articles/FrameTools.html) which at first glance appears to be a mere pretty-printer:

```{r}
library("wrapr")
```

```{r, comment=''}
cat(draw_frame(d))
```

However, the above rendering is actually executable `R` code.  If we run it, we re-create
the original `data.frame()`.

```{r}
d2 <- build_frame(
   "names", "x", "y" |
   "a"    , 1  ,  1  |
   "b"    , 2  ,  4  |
   "c"    , 3  ,  9  |
   "d"    , 4  , 16  )

print(d2)
```

The merit is: the above input is how it looks when printed.

The technique is intended for typing small examples (or [`cdata`](https://github.com/WinVector/cdata) control tables) and only builds `data.frame`s with atomic types (characters, numerics, and logicals; no times, factors or list columns).  The specification rule is the first appearance of an infix 2-argument function call (in this case the infix "or symbol" "<code>|</code>") is taken to mean the earlier arguments are part of the header or column names and later arguments are values.  The other appearances of "<code>/</code>" are ignored.  This means we could also write the frame as follows:

```{r}
build_frame(
   "names", "x", "y" |
   "a"    , 1  ,  1  ,
   "b"    , 2  ,  4  ,
   "c"    , 3  ,  9  ,
   "d"    , 4  , 16  )
```

This is more limited than `base::dump()`, but also more legible.

```{r, comment=""}
cat(dump("d", ""))
```

One can use the combination of `build_frame()` and `draw_frame()` to neaten up by-hand examples for later use (via copy and paste):

```{r, comment=""}
cat(draw_frame(build_frame(
 "names", "x", "y" |
  "a", 1,  1,
  "b", 2,  4,
  "c", 3,  9,
  "d", 4, 16)))
```

`build_frame()` allows for simple substitutions of values.  In contrast the method `qchar_frame()`
builds `data.frame`s containing only `character` types and doesn't require quoting (though it does allow it).

```{r}
qchar_frame(
  col_1, col_2, col_3 |
  a    , b    , c     |
  d    , e    , "f g" )
```

`build_frame()` is intended to capture typed-in examples, and is only compatible with very limited in-place calculation and substitution, and that _must_ be in parenthesis:

```{r}
build_frame(
   "names", "x"     , "y" |
   "a"    , 1       ,  1  |
   "b"    , cos(2)  ,  4  |
   "c"    , (3+2)   ,  9  |
   "d"    , 4       , 16  )
```

Expressions not in parenthesis (such as "<code>3 + 2</code>") will confuse the language transform `build_frame()` uses to detect cell boundaries.



