---
title: "lambda Function Builder"
author: "John Mount"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{lambda Function Builder}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The [`CRAN`](https://cran.r-project.org) version of the [`R`](https://www.r-project.org) package [`wrapr`](https://CRAN.R-project.org/package=wrapr) package now includes a concise anonymous function constructor: `λ()`.

To use it please do the following: attach `wrapr` and ask it to place a definition for `λ()`
in your environment:

```{r wrapri}
library("wrapr")
if(exists('defineLambda', 
          envir=as.environment('package:wrapr'), 
          mode='function')) {
  # Note: prior to version 0.4.2 wrapr
  # loaded a lambda-definition during
  # package load.  The following explicit 
  # wrapr::defineLambda() is more polite.
  wrapr::defineLambda() 
}
ls()
```

You can use  `λ()` to define functions.  The syntax is: `λ(arg [, arg]*, body [, env=env])`.  That
is we write a  `λ()`-call (which you can do by cutting and pasting) and list the desired function arguments and then the function body.  For example the function that squares numbers is:

```{r fsq1}
λ(x, x^2)
```

We can use such a function to square the first four positive integers as follows:

```{r fsq2}
sapply(1:4, λ(x, x^2))
```

Dot-pipe style notation does not need the `λ()` factory as it treats pipe stages
as expressions parameterized over the variable "`.`":

```{r fsqp}
1:4 %.>% {.^2}
```

And we can build functions that take more than one argument as follows:

```{r ft}
λ(x, y, x + 3*y)
```


