---
title: "Named Map Builder"
author: "John Mount"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Named Map Builder}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

"named map builder" is an operator written as "`:=`".  Named map builder is a *very* simple bit of code that performs a very simple task: it adds names to vectors or lists (making them work more like maps).

Here are some examples:

```{r ex1}
library("wrapr")

'a' := 5

c('a' := 5, 'b' := 6)

c('a', 'b') := c(5, 6)
```

The left-side argument of the `:=` operator is called "the names", and the right-side argument is called "the values".  The `:=` operators returns the values with the names set to names.

A key use of the named map builder is the following:

```{r key1}
key = 'keycode'
key := 'value'
```

Notice the value inside the variable `key` was used as the array name, this differs from 
what is easily done with `R`'s native `c(key = 'value')` style notation.


```{r print, eval=FALSE}
help(`:=`, package = 'wrapr')
```
