---
title: "Quoting Concatinate"
author: "John Mount"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quoting Concatinate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


In [`R`](https://www.r-project.org) data analysis code-capturing interfaces (or non-standard-evaluation/NSE interfaces are considered a fun convenience. These allow users to type column names in directly when working.  However in functions and packages there can be a large interface price or code safety risk to pay for supplying the user the minor convenience of eliding a few quotation marks.  

Our package [`wrapr`](https://CRAN.R-project.org/package=wrapr) has the sensible design principle: NSE can be convenient to the user, but we are going to isolate it to a few methods (for safe and simple code). Users then pass the objects that those methods create as *values* to other methods.  Computation is meant to be over values, so this is a good trade-off.

The primary place code capturing shows up in our `wrapr` `R` package is in the [`qc()`](https://winvector.github.io/wrapr/reference/qc.html) and [`qchar_frame()`](https://winvector.github.io/wrapr/reference/qchar_frame.html) methods.

`qc()` stands for "quoting concatenate" it is much like `R`'s `c()` (combine), but it quotes its arguments before concatenating vectors or lists.

It lets you replace this:

```{r}
c("Petal.Width", "Petal.Length")
```

with this:

```{r}
library("wrapr")
   
qc(Petal.Width, Petal.Length)
```



This, in turn, lets you replace this:

```
library("dplyr")

iris %>%
  select(., Petal.Width, Petal.Length) %>%
  head(.)
```

  
with this:

```{r}
iris[, qc(Petal.Width, Petal.Length)] %.>%
  head(.)
```

We still skipped the quotes, and the NSE stuff is safely isolated from the rest of the system.

`wrapr` now incorporates `bquote()` based quasiquotation in a few of its interfaces.

Quasiquotation was introduced into R by Thomas Lumley in 2003, and allows users to signal they want to turn off quotation for portions of their code. The user indicates they do not wish a portion of their code to be quoted (but instead want it evaluated for its value) by surrounding that portion with the function-notation "`.()`".

An example of this is given here.

```{r}
OTHER_SYMBOL <- "Petal.Length"
  
qc(Petal.Width, OTHER_SYMBOL)

qc(Petal.Width, .(OTHER_SYMBOL))
```

This should be familiar to `data.table` users, as `data.table` has supported related notations for quite some time.

Also, `qc()` is designed to have a simple "mutually recursive" relationship with `c()` (i.e. they call each other when they see one another).  This means `c()` is also a quasiquotation escape-notation for `qc()`:

```{r}
qc(Petal.Width, c(OTHER_SYMBOL))
```

This escape notation arises as a natural consequence of a design of `qc()` that calls `c()` instead of quoting it (i.e. delegates `c()`-expressions to `c()`). The `bquote()-.()` should be the preferred notation for regularity, and to match any other `bquote()` quasiquotation interface (such as `qchar_frame()` or [even a variation of `dplyr`](http://www.win-vector.com/blog/2018/10/quasiquotation-in-r-via-bquote/)).

`qc()` also takes some trouble to work with named vectors:

```{r}
qc(a = b)
```

And we can even re-map left-hand sizes of (or names) if we use the alternate "`:=`" assignment notation.

```{r}
LEFT_NAME = "a"

qc(.(LEFT_NAME) := b)
```

Notice syntactically `qc()` fills a general niche much like the specific function `ggplot2::aes()` fits in the `ggplot2` package.

The `qc()` notation is very powerful and clearly indicates where which quoting rules are in effect when.  We strongly suggest users look to it for code-capturing and package developers recommend it to their users.


