---
title: "Multiple Assignment with unpack"
author: "Nina Zumel and John Mount"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multiple Assignment with unpack}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In `R` there are many functions that return named lists or other structures keyed by names. Often, you want to unpack the elements of such a list into separate variables, for ease of use. One example is the use of `split()` to partition a larger data frame into a named list of smaller data frames, each corresponding to some grouping.

```{r}
library(wrapr)
# example data
d <- data.frame(
  x = 1:9,
  group = c('train', 'calibrate', 'test'),
  stringsAsFactors = FALSE)

knitr::kable(d)

# split the d by group
(parts <- split(d, d$group))
train_data <- parts$train
calibrate_data <- parts$calibrate
test_data <- parts$test

knitr::kable(train_data)

knitr::kable(calibrate_data)

knitr::kable(test_data)
```

A multiple assignment notation allows us to assign all the smaller data frames to variables in one step, and avoid leaving a possibly large temporary variable such as `parts` in our environment. One such notation is `unpack()`.

## Basic `unpack()` example

```{r}
# clear out the earlier results
rm(list = c('train_data', 'calibrate_data', 'test_data', 'parts'))

# split d and unpack the smaller data frames into separate variables
unpack(split(d, d$group),
       train_data = train,
       test_data = test,
       calibrate_data = calibrate)

knitr::kable(train_data)

knitr::kable(calibrate_data)

knitr::kable(test_data)
```

You can also use `unpack` with an assignment notation similar to the notation used with the  
<a href="https://CRAN.R-project.org/package=zeallot"><code>zeallot::%&lt;-%</code></a> pipe:

```{r}
# split d and unpack the smaller data frames into separate variables
unpack[traind = train, testd = test, cald = calibrate] <- split(d, d$group)

knitr::kable(traind)

knitr::kable(cald)

knitr::kable(testd)
```

Note: the `<-` array assignment has the minor side-effect of writing a value for `unpack` into the working environment.  This object is small, and the operation is blocked if there is an existing `unpack` object that is not such a side effect in your workspace (so it will not overwrite something else).

### Reusing the list names as variables

If you are willing to assign the elements of the list into variables with the same names, you can just use the names:

```{r}
unpack(split(d, d$group), train, test, calibrate)

knitr::kable(train)

knitr::kable(calibrate)

knitr::kable(test)

# try the unpack[]<- notation

rm(list = c('train', 'test', 'calibrate'))

unpack[test, train, calibrate] <- split(d, d$group)

knitr::kable(train)

knitr::kable(calibrate)

knitr::kable(test)
```

Mixed notation is allowed:

```{r}
rm(list = c('train', 'test', 'calibrate'))
unpack(split(d, d$group), train, holdout=test, calibrate)

knitr::kable(train)

knitr::kable(calibrate)

knitr::kable(holdout)
```


### Unpacking only parts of a list

You can also unpack only a subset of the list's elements:

```{r error=TRUE}
rm(list = c('train', 'holdout', 'calibrate'))

unpack(split(d, d$group), train, test)

knitr::kable(train)

knitr::kable(test)

# we didn't unpack the calibrate set
calibrate

```

### `unpack` checks for unknown elements

If `unpack` is asked to unpack an element it doesn't recognize, it throws an error. In this case, none of the elements are unpacked, as `unpack` is deliberately an atomic operation.

```{r error=TRUE}

# the split call will not return an element called "holdout"
unpack(split(d, d$group), training = train, testing = holdout)

# train was not unpacked either
training

```


## Other multiple assignment packages

### `zeallot`

The [`zeallot`](https://CRAN.R-project.org/package=zeallot) package already supplies excellent positional or ordered unpacking. 

The primary difference between `zeallot`'s <a href="https://CRAN.R-project.org/package=zeallot"><code>%&lt;-%</code></a> pipe and `unpack` is that `%<-%` is a *positional* unpacker: you must unpack the list based on the *order* of the elements in the list. This style may be more appropriate in the Python world where many functions return un-named tuples of results.  

`unpack` is a *named* unpacker: assignments are based on the *names* of elements in the list, and the assignments can be in any order. We feel this is more appropriate for R, as R has not emphasized positional unpacking; R functions tend to return named lists or named structures. For named lists or named structures it may not be safe to rely on value positions.

For unpacking named lists, we recommend `unpack`. For unpacking unnamed lists, use `%<-%`.

### `vadr`

<a href="https://github.com/crowding/vadr/blob/master/R/bind.R"><code>vadr::bind</code></a> supplies named unpacking, but appears to use a "<code>SOURCE = DESTINATION</code>" notation. That is the reverse of a "<code>DESTINATION = SOURCE</code>" which is how both R assignments and argument binding are already written.


