---
title: "wrapr_applicable"
author: "John Mount, Win-Vector LLC"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{wrapr_applicable}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`wrapr` includes de-referencing, function evaluation, and a new concept called `"wrapr_applicable"`.  `"wrapr_applicable"` is dispatch by type of right hand side argument scheme.

## Basic `wrapr`

The `wrapr` pipe operators (`%.>%` and `%>.%`) are roughly defined as: `a %>.% b ~ { . <- a; b };`.  This works under the assumption that `b` is an expression with free-instances of "`.`".  A typical use is:

```{r use1}
library("wrapr")

5 %.>% sin(.)
```

The above is performed by standard `S3` dispatch on the left argument of an exported generic functions called `apply_left()` and `apply_right()`.  A formal description of `wrapr` piping can be found [here](https://github.com/WinVector/wrapr/blob/master/extras/wrapr_pipe.pdf).

## Dereferencing and function evaluation

`wrapr` works primarily over expressions and "`.`".  `wrapr` does tries to de-reference names found in the right-hand side of pipe stages, and also dispatches functions.  One can also write the following.

```{r nofn}
5 %.>% sin
5 %.>% base::sin # new for wrapr 1.3.1
```

## `"wrapr_applicable"`

Arbitrary objects ask `wrapr` to treat them as special expressions by overriding one or more of `apply_left()` and `apply_right()` for the `S3` class they wish managed.

For example:

```{r sinfn}
function_reference <- list(f = sin)
class(function_reference) <- c("wrapr_applicable", "ourclass")

apply_right.ourclass <- function(pipe_left_arg,
                                 pipe_right_arg,
                                 pipe_environment,
                                 left_arg_name,
                                 pipe_string,
                                 right_arg_name) {
  pipe_right_arg$f(pipe_left_arg)
}

function_reference

5 %.>% function_reference

function_reference$f <- sqrt
5 %.>% function_reference
```

The signature arguments work as follows:

  * `pipe_left_arg`: The value moving down the pipeline.
  * `pipe_right_arg`: The right pipeline operator (essentially "`self`" or "`this`" in object oriented terms, used for `S3` dispatch).
  * `pipe_environment`: The environment the pipeline is working in (not usually needed).
  * `left_arg_name`: If the left arguement was passed in by name, what that name was.
  * `pipe_string`: The name of the pipe operator (not usually needed).
  * `right_arg_name`: If the right arguement was passed in by name, what that name was.


This functionality allows arbitrary objects to directly specify their intended pipeline behavior.

Let's use a debugging function to see the values of all of the arguments.

```{r debug}
apply_right.ourclass <- function(pipe_left_arg,
                                 pipe_right_arg,
                                 pipe_environment,
                                 left_arg_name,
                                 pipe_string,
                                 right_arg_name) {
  print("pipe_left_arg")
  print(pipe_left_arg)
  print("pipe_right_arg")
  print(pipe_right_arg)
  print("pipe_environment")
  print(pipe_environment)
  print("left_arg_name")
  print(left_arg_name)
  print("pipe_string")
  print(pipe_string)
  print("right_arg_name")
  print(right_arg_name)
  pipe_right_arg$f(pipe_left_arg)
}

5 %.>% function_reference

a <- 5

a %.>% function_reference
```



## Conclusion

`wrapr` values (left-hand sides of pipe expressions) are completely general.
`wrapr` operators (right-hand sides of pipe expressions) are primarily intended to be expressions that have "`.`" as a free-reference.  `wrapr` can also be used with right-hand sides that are function references or with arbitrary annotated objects.


