---
title: "wrapr_applicable"
author: "John Mount, Win-Vector LLC"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{wrapr_applicable}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`wrapr` `1.1.0` includes de-referencing, function evaluation, and a new concept called `"wrapr_applicable"`.

## Basic `wrapr`

The `wrapr` pipe operators (`%.>%` and `%>.%`) are roughly defined as: `a %>.% b ~ { . <- a; b };`.  This works under the assumption that `b` is an expression with free-instances of "`.`".  A typical use is:

```{r use1}
library("wrapr")

5 %.>% sin(.)
```

## Dereferencing and function evaluation

`wrapr` works primarily over expressions and "`.`".  As of version `1.1.0` `wrapr` does try to de-reference names found in the right-hand side of pipe stages, and also dispatches functions.  That is: one can now write the following.

```{r nofn}
5 %.>% sin
```

## `"wrapr_applicable"`

Arbitrary objects ask `wrapr` to treat them as special expressions by:

  1. Adding the string `"wrapr_applicable"` to their class declarations.
  2. Populating an list entry named `"wrapr_function"` with a function of signature `(pipe_left_arg, pipe_right_arg, pipe_environment)`.
  
If these two conditions are met then `wrapr` instead of evaluating the object calls the function `wrapr_function`.

For example:

```{r sinfn}
function_reference <- list(f = sin)
class(function_reference) <- "wrapr_applicable"
function_reference$wrapr_function <- function(pipe_left_arg, 
                                              pipe_right_arg, 
                                              pipe_environment) {
  pipe_right_arg$f(pipe_left_arg)
}

function_reference

5 %.>% function_reference

function_reference$f <- sqrt
5 %.>% function_reference
```

The signature arguments work as follows:

  * `pipe_left_arg`: The value moving down the pipeline.
  * `pipe_right_arg`: The right pipeline operator (essentially "`self`" or "`this`" in object oriented terms).
  * `pipe_environment`: The environment the pipeline is working in (not usually needed).

This functionality allows arbitrary objects to directly specify their intended pipeline behavior.

## Application

[`rquery`](https://github.com/WinVector/rquery) is a first application of this new `"wrapr_applicable"` feature.  This [allows `rquery` pipelines to be applied to directly to `data.frame`s](https://winvector.github.io/rquery/reference/rquery_apply_to_data_frame.html) (even though `rquery` is primarily intended for databases).  A discussion of the methodology can be found [here](https://winvector.github.io/rquery/articles/AdHocQueries.html#stored-ad-hoc-pipeline-mode).


## Conclusion

`wrapr` values (left-hand sides of pipe expressions) are completely general.
`wrapr` operators (right-hand sides of pipe expressions) are primarily intended to be expressions that have "`.`" as a free-reference.  `wrapr` can also be used with right-hand sides that are function references or with arbitrary annotated objects.


