---
title: "Piple Performance"
author: "John Mount"
date: "7/7/2017"
output:
  md_document:
    variant: markdown_github
---

Some timings for [`%.>%`](http://www.win-vector.com/blog/2017/07/in-praise-of-syntactic-sugar/).

Keep in mind for any *serious* application the calculation time on data will far dominate any piping overhead, but it is fun to look.

So we will compare:

 * `magrittr*` `magrittr::%>%` substitution.
 * `DotBlockPipe*` `wrapr::%.>%` substitution.
 * `BizarroPipe*` `->.;` substitution.
 
```{r generator, include=FALSE, echo=FALSE}
# build source code for fns of size k
# build source code for fns of size k
buildFnsK <- function(k) {
  fs <- c("5", rep("sin(.)", k))
  fMap <- list('magrittr' = '%>%' , 
               'DotBlockPipe' = '%.>%', 
               'BizarroPipe' = '->.;')
  fns  <- lapply(names(fMap),
                 function(fn) {
                   op <- fMap[[fn]]
                   paste0(fn, '_', k, ' <- function() {\n ',
                          paste(fs, collapse = paste0(' ', op, '\n   ')),
                          '\n}\n\n')
                   
                 })
  paste(fns, collapse= '\n\n')
}


ftext <- paste(
  buildFnsK(1),
  buildFnsK(2),
  buildFnsK(5),
  buildFnsK(10),
  buildFnsK(15),
  buildFnsK(20),
  buildFnsK(25),
  buildFnsK(50),
  collapse = "\n\n")

fileConn<-file("pGenFns.R")
writeLines(ftext, fileConn)
close(fileConn)
```

```{r defs}
library("microbenchmark")
library("wrapr")
library("rlang")
suppressPackageStartupMessages(library("ggplot2"))
suppressPackageStartupMessages(library("dplyr"))

# load generated examples
source("pGenFns.R")

print(BizarroPipe_5)

print(DotBlockPipe_5)

print(magrittr_5)

BizarroPipe_10()

DotBlockPipe_10()

magrittr_10()
```

```{r timings}
bm <- microbenchmark(
  magrittr_1(),
  DotBlockPipe_1(),
  BizarroPipe_1(),
  magrittr_2(),
  DotBlockPipe_2(),
  BizarroPipe_2(),
  magrittr_5(),
  DotBlockPipe_5(),
  BizarroPipe_5(),
  magrittr_10(),
  DotBlockPipe_10(),
  BizarroPipe_10(),
  magrittr_15(),
  DotBlockPipe_15(),
  BizarroPipe_15(),
  magrittr_20(),
  DotBlockPipe_20(),
  BizarroPipe_20(),
  magrittr_25(),
  DotBlockPipe_25(),
  BizarroPipe_25(), 
  magrittr_50(),
  DotBlockPipe_50(),
  BizarroPipe_50(), 
  times=1000L
)
print(bm)
autoplot(bm)
```

```{r replot}
d <- as.data.frame(bm)
d$size <- as.numeric(gsub("[^0-9]+", "", d$expr))
d$fn <- gsub("[_0-9].*$", "", d$expr)

mkPlot <- function(d, title) {
  d$size <- as.factor(d$size)
  highCut <- as.numeric(quantile(d$time, probs = 0.99))
  dcut <- d[d$time<=highCut, , drop=FALSE]
  
  ggplot(data=dcut, aes(x=time, group=expr, color=size)) +
    geom_density(adjust=0.3) +
    facet_wrap(~fn, ncol=1, scales = 'free_y') +
    xlab('time (NS)') + ggtitle(title)
}

mkPlot(d, 'all timings')
mkPlot(d[d$fn %in% c('magrittr', 'DotBlockPipe'), , drop=FALSE], 
       'magrittr v.s. DotBlockPipe')
```

```{r calculate}
# fit a linear function for runtime as a function of size
# per group.
fits <- d %.>%
  split(., .$fn) %.>%
  lapply(., 
         function(di) { 
           lm(time ~ size, data=di) 
         }) %.>%
  lapply(., coefficients) %.>%
  lapply(., 
         function(ri) {
           data.frame(Intercept= ri[["(Intercept)"]],
                      size= ri[['size']])
         }) 
dfits <- bind_rows(fits)
dfits$fn <- names(fits)

# "Intercept" is roughly start-up cost 
# "size" is slope or growth rate
print(dfits)

# solve for size where two lines interesect.
# Note: this is a naive estimate, and not stable
# in the face of estimated slopes and intercepts.
solve <- function(dfits, f1, f2) {
  idx1 <- which(dfits$fn==f1)
  idx2 <- which(dfits$fn==f2)
  size <- (dfits$Intercept[[idx1]] - dfits$Intercept[[idx2]]) /
    (dfits$size[[idx2]] - dfits$size[[idx1]])
  size
}

crossingPoint <- solve(dfits, 'DotBlockPipe', 'magrittr')
print(crossingPoint)

ratio <- dfits$size[dfits$fn=='DotBlockPipe'] / dfits$size[dfits$fn=='magrittr']
print(ratio)
```

Overall:

DotBlockPipe is about `r format(ratio, digits=2)` times slower than magrittr on large pipelines, though this cost will be unnoticeable with in any significant workload.

