---
title: "Starting with wrapr::let()"
author: "John Mount"
date: "`r Sys.Date()`"
output:
  md_document:
    variant: markdown_github
---


#  `wrapr::let()` for beginners

`wrapr::let()` is not interesting if you have no idea why you would need it or how to use it.

`wrapr::let()` is designed to solve a single problem when programming in `R`: when in writing code you don't yet know the name of a variable or a `data.frame` column you are going to use. That is: the name of the column will only be available later when your code is run.

For example suppose we have a `data.frame` "`d`" defined as follows:

```{r data}
d <- data.frame(x = c(15, 30, 40))
print( d )
```

If we know the name of the column we can access it as follows:

```{r std} 
print( d$x )
```

If we don't know the name of the column (such as would be the case when writing a function) we write code like the following:

```{r param}
getColumn <- function(df, columnName) {
  df[[columnName]]
}

print( getColumn(d, 'x') )
```

This works because `R` takes a lot of trouble to supply parametric interfaces for most use cases.

We only run into trouble if code we are trying to work with strongly prefers non-parametric (also called non-standard-evaluation or NSE) interfaces. 

A popular package that heavily emphasizes NSE interfaces is `dplyr`.  It is the case that `dplyr` supplies its own methods to work around NSE issues ("underbar methods" and `lazyeval` for `dplyr` `0.5.0`, and `rlang`/`tidyeval` for `dplyr` `0.7.0`).  However, we will only discuss `wrapr::let()` methodology here.

The issue is: it is common to write statements such as the following in `dplyr`:

```{r dplyr1}
suppressPackageStartupMessages(library("dplyr"))

d %>% mutate( v2 = x + 1 )
```

If we were writing the above in a function it would plausible that we would not know the name of the desired result column or the name of the column to add one to.  `wrapr::let()` lets us write such code easily:

```{r let1}
library("wrapr")

addOneToColumn <- function(df, 
                           result_column_name, 
                           input_column_name) {
  let(
    c(RESULT_COL = result_column_name,
      INPUT_COL = input_column_name),
    
    df %>% mutate( RESULT_COL = INPUT_COL + 1 )
    
  )
}

d %>% addOneToColumn('v2', 'x')
```

Again, writing the function `addOneToColumn()` was the goal.  The issue was that in such a function we don't know what column names the user is going to end up supplying.  We work around this difficulty with `wrapr::let()`  

All the user needs to keep in mind is: `wrapr::let()` takes two primary arguments:

 * The aliases mapping concrete stand-in names (here shown in `ALL_CAPS` style) to the variables holding the actual variable names.
 * The block of code to re-write and execute (which can be a single statement, or a larger block with braces).

