---
title: "Reusable Pipelines in R"
output: github_document
---


This note will discuss the [`R`](https://www.r-project.org) features: [`rquery`](https://github.com/WinVector/rquery)/[`rqdatatable`](https://github.com/WinVector/rqdatatable) operator trees and [`wrapr` function object pipelines](https://winvector.github.io/wrapr/articles/Function_Objects.html).

For quite a while the [`rquery`](https://github.com/WinVector/rquery) and [`rqdatatable`](https://github.com/WinVector/rqdatatable) packages have supplied a sequence of operators abstraction called an "operator tree" or "operator pipeline".  

These pipelines are (deliberately) fairly strict:

  * They must start with a table description or definition.
  * Each step must be a table to table transform meeting certain column pre-conditions.
  * Each step must advertise what columns it makes available or produces, for later condition checking.
  
For our example suppose we want to subset some data, get per-group means, and then sort the data by those means.

```{r}
# our example data
d <- data.frame(
  group = c("a", "a", "b", "b"),
  value = c(  1,  2,   2,  -10),
  stringsAsFactors = FALSE
)

# load our package
library("rqdatatable")

# build an operator tree
threshold <- 0.0
ops <-
  # define the data format
  local_td(d) %.>%   
  # restrict to rows with value >= threshold
  select_rows_nse(.,
                  value >= threshold) %.>%
  # compute per-group aggegations
  project_nse(.,
              groupby = "group",
              mean_value = mean(value)) %.>%
  # sort rows by mean_value decreasing
  orderby(.,
          cols = "mean_value",
          reverse = "mean_value")

# show the tree/pipeline
cat(format(ops))
```

Of course the purpose of such a pipeline is to be able to apply it to data.  This is done simply with the [`wrapr` dot arrow pipe](https://journal.r-project.org/archive/2018/RJ-2018-042/index.html):

```{r}
d %.>% ops
```

`rquery` pipelines are designed to specify and execute data wrangling tasks. An important feature of `rquery` pipelines is: they are designed for serialization.  This means we can save them and also send them to multiple nodes for parallel processing.


```{r}
# save the optree
saveRDS(ops, "rquery_optree.RDS")

# simulate a fresh R session
rm(list=setdiff(ls(), "d"))

library("rqdatatable")

# read the optree back in
ops <- readRDS('rquery_optree.RDS')

# look at it
cat(format(ops))

# use it again
d %.>% ops

# clean up
rm(list=setdiff(ls(), "d"))
```

We can also run `rqdatatable` operations in "immediate mode", without pre-defining the pipeline or tables:

```{r}
threshold <- 0.0

d %.>%
  select_rows_nse(.,
                  value >= threshold) %.>%
  project_nse(.,
              groupby = "group",
              mean_value = mean(value)) %.>%
  orderby(.,
          cols = "mean_value",
          reverse = "mean_value")
```

A natural question is: given we already have `rquery` pipelines why do we need [`wrapr` function object pipelines](https://winvector.github.io/wrapr/articles/Function_Objects.html)? The reason is: `rquery`/`rdatatable` pipelines are strict and deliberately restricted to operations that can be hosted both in `R` (via `data.table`) or on databases (examples: `PostgreSQL` and `Spark`).  One might also want a more general pipeline with fewer constraints optimized for working in `R` directly.  

The `wrapr` "function object" pipelines allow treatment of arbitrary objects as items we can pipe into.  Their primary purpose is to [partially apply functions](https://en.wikipedia.org/wiki/Partial_application) to convert arbitrary objects and functions into single-argument (or unary) functions.  This converted form is perfect for pipelining. This, in a sense, lets us treat these objects as functions.  The `wrapr` function object pipeline also has less constraint checking than `rquery` pipelines, so is more suitable for "black box" steps that do not publish their column use and production details (in fact `wrapr` function object pipelines work on arbitrary objects, not just `data.frame`s or tables).

Let's adapt our above example into a simple `wrapr` dot arrow pipeline.


```{r}
library("wrapr")

threshold <- 0

d %.>%
  .[.$value >= threshold, , drop = FALSE] %.>%
  tapply(.$value, .$group, 'mean') %.>%
  sort(., decreasing = TRUE)
```

All we have done is replace the `rquery` steps with typical base-`R` commands. As we see the `wrapr` dot arrow can route data through a sequence of such commands to repeat our example.

Now let's adapt our above example into a re-usable `wrapr` function object pipeline.

```{r}
library("wrapr")

threshold <- 0

pipeline <-
  srcfn(
    ".[.$value >= threshold, , drop = FALSE]" ) %.>%
  srcfn(
    "tapply(.$value, .$group, 'mean')" ) %.>%
  pkgfn(
    "sort",
    arg_name = "x",
    args = list(decreasing = TRUE))

cat(format(pipeline))
```


We used two `wrapr` abstractions to capture the steps for re-use (something built in to `rquery`, and now also supplied by `wrapr`).  The abstractions are:

  * [`srcfn()`](https://winvector.github.io/wrapr/reference/srcfn.html) which wraps arbitrary quoted code as a function object.
  * [`pkgfn()`](https://winvector.github.io/wrapr/reference/pkgfn.html) which wraps a package qualified function name as a function object ("`base`" being the default package).
  
This sort of `pipeline` can be applied to data using pipe notation:


```{r}
d %.>% pipeline
```

The above pipeline has one key inconvenience and one key weakness:

  * For the `srcfn()` steps we had to place the source code in quotes, which defeats any sort of syntax highlighting and auto-completing in our `R` integrated development environment (IDE).
  * The above pipeline has a reference to the value of `threshold` in our current environment, this means the pipeline is not sufficiently self-contained to serialize and share.
  
We can quickly address both of these issues with the [`wrapr::qe()`](https://winvector.github.io/wrapr/reference/qe.html) ("quote expression") function.  It uses `base::substitute()` to quote its arguments, and the IDE doesn't know the contents are quoted and thus can help us with syntax highlighting and auto-completion.  Also we are using [`base::bquote()` .()-style](http://www.win-vector.com/blog/2018/09/parameterizing-with-bquote/) escaping to bind in the value of `threshold`.

```{r}
pipeline <-
  srcfn(
    qe( .[.$value >= .(threshold), , drop = FALSE] )) %.>%
  srcfn(
    qe( tapply(.$value, .$group, 'mean') ))  %.>%
  pkgfn(
    "sort",
    arg_name = "x",
    args = list(decreasing = TRUE))

cat(format(pipeline))

d %.>% pipeline
```

Notice this pipeline works as before, but no longer refers to the external value `threshold`.

A recommended way to bind in values is with the `args`-argument, which is a named list of values that are expected to be available with a `srcfn()` is evaluated, or additional named arguments that will be applied to a `pkgfn()`.

In this notation the pipeline is written as follows.

```{r}
pipeline <-
  srcfn(
    qe( .[.$value >= threshold, , drop = FALSE] ),
    args = list('threshold' = threshold)) %.>%
  srcfn(
    qe( tapply(.$value, .$group, 'mean') ))  %.>%
  pkgfn(
    "sort",
    arg_name = "x",
    args = list(decreasing = TRUE))

cat(format(pipeline))

d %.>% pipeline
```


We can save this pipeline.


```{r}
saveRDS(pipeline, "wrapr_pipeline.RDS")
```

And simulate using it in a fresh environment (i.e. simulate sharing it).


```{r}
# simulate a fresh environment
rm(list = setdiff(ls(), "d"))

library("wrapr")

pipeline <- readRDS('wrapr_pipeline.RDS')

cat(format(pipeline))

d %.>% pipeline
```

And that is some of the power of `wrapr` function objects.  This may seem laborious when the steps are simple transforms, however the technique is very convenient when each of the steps is a substantial (such as non-trivial data preparation and model application steps). 

The above techniques can make reproducing and sharing methods much easier.

We have some more examples of the technique [here](http://www.win-vector.com/blog/2018/12/sharing-modeling-pipelines-in-r/) and [here](https://winvector.github.io/wrapr/articles/Function_Objects.html).


```{r}
# clean up after example
unlink("rquery_optree.RDS")
unlink("wrapr_pipeline.RDS")
```
