---
title: "Some wrapr Notations"
author: "Win-Vector LLC"
date: "1/21/2018"
output: github_document
---

I would like to demonstrate some helpful [`wrapr`](https://CRAN.R-project.org/package=wrapr) [`R`](https://www.r-project.org) notation tools that really neaten up your `R` code.

## Named Map Builder

First we demonstrate `wrapr`'s ["named map builder": `:=`](https://winvector.github.io/wrapr/reference/named_map_builder.html).  
The named name builder adds names to vectors and lists by nice "names on the left and values on the right" notation.

For example to build a named vector mapping names `c("a", "b")` to values `c(1, 2)` we could write the following `R` code.

```{r br1}
c("a" = 1, "b" = 2)
```

Using  `wrapr` we can write the same thing using `:=`.

```{r nmb1}
library("data.table") # data.table before wrapr to avoid := contention
suppressPackageStartupMessages(library("dplyr"))
library("wrapr")


c("a" := 1, "b" := 2)
```

This is read as "`a` is 1 and `b` is 2".

So far, no gain. However (unlike `<-`), `:=` also works vectorized (as shown below).

```{r nmb2}
c("a", "b") := c(1, 2)
```

One can think of this as an operator version of `setNames(nm = c("a", "b"), c(1,2))` (from `stats`).  This notation is very handy once you look for places to use it and for tools to further speed it up.

### Quoting Combine

[`qc()` (quoting combine/concatenate)](https://winvector.github.io/wrapr/reference/qc.html) is another wrapr notation improving function.  `qc()` work by analogy
to `R`'s `c()` function, except it quotes its arguments.  This lets us write the previous vector naming as:

```{r nmb3}
qc(a, b) := c(1, 2)
```

## `let()`

The above notations work particularly well with `wrapr::let()`. 

`wrapr::let()` evaluates an expression (or block of expressions) with a number of symbolic substitutions.  The named map builder is a great way to specify such substitutions.  A quick examples is computing "1 plus variable" where the actual name of the variable is specified in a named vector called "`mapping`".

```{r let1}
mapping <- "VARNAME" := "x"
x <- 1 # the actual varaible

# Evaluate "VARNAME + 1" with mapping substitution
let(alias = mapping,
    VARNAME + 1
    )
```

The mapping from abstract variables (variable names used in the code) to concrete variables (variable names with desired values in the execution environment) can also be performed inline:

```{r let1b}
let("VARNAME" := "x",
    VARNAME + 1
    )
```

`let()` itself is useful in writing re-usable (or parametric) functions (often a challenge in `R`).

```{r let2}

d <- data.frame(x = c(1, 2))

incrementColumn <- function(data, COLUMNNAME) {
  let(c("COLUMNNAME" = COLUMNNAME),
      mutate(data, COLUMNNAME = COLUMNNAME + 1)
  )
}

incrementColumn(d, "x")
```

The idea is we use the stand-in symbol `COLUMNNAME` in our code (no matter how complicated) and
`let()` substitutes name (represented as a string) stored into the expression before execution.
What we just executed was equivalent to the following:

```{r let3}
COLUMNNAME = "x"
let(c("COLUMNNAME" = COLUMNNAME), eval = FALSE,
      mutate(d, COLUMNNAME = COLUMNNAME + 1)
  )
```


### `mapsyms()` (the `let(X=X)`, replace with value, convention)

[`wrapr::mapsyms()`](https://winvector.github.io/wrapr/reference/mapsyms.html) is a helper function makes function creation even more convenient.  A 
`mapsyms` expression of the form `mapsyms(COLUMNNAME)` is equivalent to the code `c("COLUMNNAME" = COLUMNNAME)`.
In our example that means it builds the name to name mapping: `r map_to_char(mapsyms(COLUMNNAME))` (here we used [`wrapr::map_to_char()`](https://winvector.github.io/wrapr/reference/map_to_char.html) to present the result).  With 
`mapsyms()` we can write the earlier function as:


```{r increxfn}
incrementColumn <- function(data, COLUMNNAME) {
  let(mapsyms(COLUMNNAME),
      mutate(data, COLUMNNAME = COLUMNNAME + 1)
  )
}

incrementColumn(d, "x")
```

We have more `mapsyms()` examples in our article ["Let X=X in R"](http://www.win-vector.com/blog/2017/11/let-xx-in-r/).

The `let()` method of building functions works well with [`dplyr`](https://CRAN.R-project.org/package=dplyr) and [`data.table`](https://CRAN.R-project.org/package=data.table).  For each of these let's show 
code for the "by hand logistic scoring" example from ["Let’s Have Some Sympathy For The Part-time R User"](http://www.win-vector.com/blog/2017/08/lets-have-some-sympathy-for-the-part-time-r-user/).

The `dplyr` logistic scoring function example is as follows.

```{r dplyr1}
# scoring function wrapping code where the 
# columns of interest were originally
# subjectID, surveyCategory, and assessmentTotal.
# function lets us re-map columns to new data later.
logistic_score <- function(data, scale, 
                           subjectID = "subjectID",
                           surveyCategory = "surveyCategory",
                           assessmentTotal = "assessmentTotal") {
  let(mapsyms(subjectID, surveyCategory, assessmentTotal),
      data %>%
        group_by(subjectID) %>%
        mutate(probability =
                 exp(assessmentTotal * scale)/
                 sum(exp(assessmentTotal * scale))) %>%
        arrange(probability, surveyCategory) %>%
        mutate(isDiagnosis = row_number() == n()) %>%
        filter(isDiagnosis) %>%
        ungroup() %>%
        select(subjectID, surveyCategory, probability) %>%
        rename(diagnosis = surveyCategory) %>%
        arrange(subjectID)
      )
}

# new data where columns of interest are
# student, surveyCategory, and points.
d <- data.frame(
  student = c(1, 1, 2, 2),
  surveyCategory = c(
    'withdrawal behavior', 'positive re-framing',
    'withdrawal behavior', 'positive re-framing'
  ),
  points = c(5, 2, 3, 4),
  stringsAsFactors = FALSE
)

# apply the function to new data
# re-specifying columns to the names we need
d %>%
  logistic_score(scale = 0.237,
                 subjectID = "student",
                 assessmentTotal = "points") %>%
  knitr::kable()
```

### The let(X=x) (mixed case) convention

For cases where the original code already has a mixture of parametric 
specifications (column names taken from variables) and non-parametric 
specifications (column names captured from un-evaluated code) we
suggest using the "mixed case" convention.  In mixed case convention
all upper case symbols are used for replacement and lower case are taken as values.
This is just a convention (the code does not implement the above as a rule)
and we specify it by forming let alias maps of the form `qc(X=x)` which means
in the `let`-block any instances of `X` are replaced with the name stored in `x`
and (naturally) any instances of `x` are left alone.

Hers is a `data.table` function example  ([from here](http://www.win-vector.com/blog/2018/01/base-r-can-be-fast/#comment-66751)) using the mixed case convention.

```{r dt1}
logistic_score <- function(data, scale, 
                           subjectID = "subjectID",
                           surveyCategory = "surveyCategory",
                           assessmentTotal = "assessmentTotal") {
  let(qc(SUBJECTID, SURVEYCATEGORY, ASSESSMENTTOTAL) :=
        c(subjectID, surveyCategory, assessmentTotal),
      { 
        dDT <- data.table::data.table(data)
        setnames(dDT, surveyCategory, "diagnosis")
        dDT[, expaTs := exp(ASSESSMENTTOTAL * scale)]
        # precalculate -> this uses gsum internally
        dDT[, sum_expaTs := sum(expaTs), SUBJECTID]
        dDT[, probability := expaTs / sum_expaTs]
        dDT[, c(assessmentTotal, "expaTs", "sum_expaTs") := NULL]
        setorder(dDT, SUBJECTID,-probability, diagnosis)
        dDT[, .SD[1], SUBJECTID]
      })
}

d %>%
  logistic_score(., 
                 scale = 0.237,
                 subjectID = "student",
                 assessmentTotal = "points") %>%
  knitr::kable(.)
```

To use the mixed case convention:

  * Build a name map using the `qc(...) := c(..)` notation. Reserve uppercase symbols for symbols to be replaced and lower case symbols for variable holding names of columns.  
  * Given the above use the uppercase version of each symbol in non-standard contexts, and lower-case in standard contexts.
  
The mixed case convention is *very* powerful.

The `1.1.2` version of `wrapr` adds a new function `map_upper()` which allows writing the `qc(SUBJECTID, SURVEYCATEGORY, ASSESSMENTTOTAL) := c(subjectID, surveyCategory, assessmentTotal)` simply as `map_upper(subjectID, surveyCategory, assessmentTotal)`:

```{r mu}
subjectID = "student"
surveyCategory = "surveyCategory"
assessmentTotal = "points"

map_upper(subjectID, surveyCategory, assessmentTotal)
```

I would like to call out that all of these `wrapr` features (`:=`, `qc()`, `mapsyms()` `map_upper()`, `let()`) are concrete functions that can be used separately or used together.  That is: `:=` isn't a symbol that has a new interpretation only in `let()` blocks, it is a inline function that actually builds named vectors, and these named vectors in turn happen to be able to specify the mappings `let()` needs. This allows you to learn and test these functions separately (and allows you to find new uses for them in your own code).  For example: if you find a new way to use `let()` blocks that needs a new mapping function, you can build that function (as the current functions are not wired into `let()`, so are not magic or privileged).

For multi-expression `let()`-blocks we must add `{}`. For `:=` to work we must have `wrapr`'s definition active, which we achieved by loading the `wrapr` package after loading the `data.table` package.  `data.table`'s use of `:=` should continue to be correct as that is always performed by `data.table` itself, where `wrapr`'s definition can not interfere.

# Take Away

`wrapr` supplies some powerful and convenient `R` notations. I hope you can incorporate `wrapr` into your work, and please do check out some of our additional training materials.


